A Mechanised Library for Generic (Co)Programming

### 2024-02-12 paper/paper.tex
Some cleanup about containers and container extensions

A container is defined by a type of \emph{shapes} (\mintinline{Coq}{Shape }),
and a \emph{family of position types} indexed by shapes
(\mintinline{Coq}{Pos}).  An \emph{extension} of this container is a functor,
whose action on objects is given by the following definition:
\begin{minted}{Coq}
  Record App (Shape : Set) (Pos : Shape -> Set) (X : Set) :=
    { shape : Shape; contents : Pos shape -> X }.
\end{minted}
In our mechanisation, we use \mintinline{Coq}{App} to represent container
extensions.  Generally, any functor that is isomorphic to a container extension
is called a \emph{container functor}. However, in our work we will call
\emph{container functor} to the functors defined by a specific container
extension.

The first problem in extracting Coq code from such container extensions is the
use of families of types. Since OCaml's type system cannot accurately capture
such families of types, Coq's code extraction mechanism will insert unsafe
casts. To avoid unsafe casts, we can consider representing the positions of a
container as a type that captures \emph{all} the potential positions in all
possible shapes, together with a predicate \mintinline{Coq}{Valid} that
determines whether a position is valid in a shape. The family of positions is
then equivalent to the type \mintinline{Coq}{fun shape => {p | Valid shape p}}.
\begin{minted}{Coq}
  Record App (Shape : Set) (Pos : Set)
                   (Valid : Shape -> Pos -> Prop) (X : Set) :=
    { shape : Shape; contents : {p | Valid shape p} -> X }.
\end{minted}
This approach solves the the problem of the unsafe casts, since the code
extracted for \mintinline{Coq}{{p | Valid shape pos}} will be an OCaml
singleton type defined as the OCaml equivalent to \mintinline{Coq}{Pos}.

However, there still remains the issue of reasoning about the equality of
container extensions. The contents of the container extensions need to be
accessed via a function \mintinline{Coq}{contents}. The domain of this function
depends on the shape of the container. This means that, for a given type of
shapes and positions, the equality of two objects of a container extension will
need to deal with heterogeneous equality. Since we want to remain axiom-free,
one way to deal with this is to rely on a \emph{decidable} validity predicate.
This way, we reach our final definition for container extensions. First,
containers are defined as pairs of shape and position types that have a
\emph{decidable} validity predicate. We represent decidable predicates as
functions that return a boolean, and lift booleans to propositions similarly to
SSReflect.
\begin{minted}{Coq}
Class Container (Shape : Set) (Pos : Set) := {
   valid : Shape -> Pos -> bool
}.
Coercion is_true (b : bool) : Prop := b = true.
\end{minted}
Then, we define \emph{container extensions} as
\begin{minted}{Coq}
  Record App `{container Shape Pos} (X : Set) :=
    { shape : Shape; contents : {p | valid shape pos} -> X }.
\end{minted}

>>>>

A container is defined by a type of \emph{shapes}, and a \emph{family of
position types} indexed by shapes. 
\begin{minted}{Coq}
Context (Shape : Set) (Pos : Shape -> Set).
\end{minted}
An \emph{extension} of this container is a functor,
whose action on objects is given by:
\begin{minted}{Coq}
Record App (X : Set) := { shape : Shape; contents : Pos shape -> X }.
\end{minted}
In our mechanisation, we use \mintinline{Coq}{App} to represent container
extensions.  Generally, any functor that is isomorphic to a container extension
is called a \emph{container functor}. However, in our work we will call
\emph{container functor} to the functors defined by a specific container
extension.

The first problem in extracting Coq code from such container extensions is the
use of families of types. Since OCaml's type system cannot accurately capture
such families of types, Coq's code extraction mechanism will insert unsafe
casts. To avoid unsafe casts, we can consider representing the positions of a
container as a type that captures \emph{all} the potential positions in all
possible shapes, together with a predicate \mintinline{Coq}{Valid} that
determines whether a position is valid in a shape. 
\begin{minted}{Coq}
Context (Shape : Set) (Pos : Set) (Valid : Shape -> Pos -> Prop).
Record App (X : Set) := { shape    : Shape; 
                          contents : {p | Valid shape p} -> X }.
\end{minted}
This approach solves the the problem of the unsafe casts, since the code
extracted for \mintinline{Coq}{{p | Valid shape pos}} will be an OCaml
singleton type defined as the OCaml equivalent to \mintinline{Coq}{Pos}.

However, there still remains the issue of reasoning about the equality of
container extensions. The contents of the container extensions need to be
accessed via a function \mintinline{Coq}{contents}. The domain of this function
depends on the shape of the container. This means that, for a given type of
shapes and positions, the equality of two objects of a container extension will
need to deal with heterogeneous equality. Since we want to remain axiom-free,
one way to deal with this is to rely on a \emph{decidable} validity predicate.
This way, we reach our final definition for container extensions. First,
containers are defined as pairs of shape and position types that have a
\emph{decidable} validity predicate. We represent decidable predicates as
functions that return a boolean, and lift booleans to propositions similarly to
SSReflect.
\begin{minted}{Coq}
Class Cont (Shape : Set) (Pos : Set) := { valid : Shape -> Pos -> bool }.
Coercion is_true (b : bool) : Prop := b = true.
\end{minted}
Then, we define \emph{container extensions} as
\begin{minted}{Coq}
Record App `(C : Cont Shape Pos) (X : Set) :=
  { shape : Shape; contents : {p | valid shape pos} -> X }.
\end{minted}

The functorial action for container extensions on morphisms boils down to
function composition, which trivially preserves identity and composition.
\begin{minted}{Coq}
Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
  := {| shape := shape x; contents := fun e => f (contents x e) |}
\end{minted}

\subsubsection{Polynomial Functors as Container Extensions}

\subsubsection{Equality of Container Extensions}
